---
title: "Adaptive Walks"
output: html_notebook
author: Ted Monyak
description: This notebook contains scripts for understanding the dynamics of adaptive walks.
---
Imports
```{r}
library(AlphaSimR)
library(devtools)
library(dplyr)
library(ggplot2)
library(factoextra)
library(patchwork)
library(plotly)
library(purrr)
library(reshape2)
library(snow)
library(tibble)
library(tidyr)
library(viridis)
library(notly)
rm(list = ls())

set.seed(123)

source("Functions/Fitness.R")
source("Functions/TraitArchitecture.R")
source("Scripts/GlobalParameters.R")
#source("Scripts/CreateFounderPop.R")

# If running on Alpine
setwd("/pl/active/Morris_CSU/Ted_Monyak/BreedingSims")
output_dir <- ("/scratch/alpine/c837220672@colostate.edu/Output")

# If running locally:
#setwd("~/Documents/CSU/R/BreedingSims")
#output_dir <- file.path(getwd(), "Output")

if (!dir.exists(output_dir)) dir.create(output_dir)
```


```{r setup, include=FALSE, echo=FALSE}
require("knitr")
# If running locally:
opts_knit$set(root.dir = "~/Documents/CSU/R/BreedingSims")

# If running on alpine:
opts_knit$set(root.dir = "/pl/active/Morris_CSU/Ted_Monyak/BreedingSims")
```
```

View a fitness landscape
```{r}
p <- plotFitnessLandscape()

save_dir <- file.path(output_dir, "FitnessFunctions")
if (!dir.exists(save_dir)) dir.create(save_dir)
save_dir <- file.path(save_dir, format(Sys.time(), "%F_%H_%M_%S"))
if (!dir.exists(save_dir)) dir.create(save_dir)
fname <- file.path(save_dir, "fitness_function.html")
htmlwidgets::saveWidget(as_widget(p), fname)
```

Plot different trait architectures according to different algorithms
```{r}
save_dir <- file.path(output_dir, "TraitArchitecture")
if (!dir.exists(save_dir)) dir.create(save_dir)
save_dir <- file.path(save_dir, format(Sys.time(), "%F_%H_%M_%S"))
if (!dir.exists(save_dir)) dir.create(save_dir)
write.table(getParams(), file.path(save_dir, "params.txt"), col.names=FALSE, quote=FALSE, sep=":\t")

# Additive effects
methodType <- "Additive"
g <- plotTraitArchitecture(founderPop, methodType)
ggplot2::ggsave(filename = paste0("traitarchitecture_", methodType, ".pdf"),
                path=save_dir,
                device = "pdf")

# Effect of each allele on fitness
methodType <- "Fitness"
g <- plotTraitArchitecture(founderPop, methodType)
ggplot2::ggsave(filename = paste0("traitarchitecture_", methodType, ".pdf"),
                path=save_dir,
                device = "pdf")

p <- plot3dPopulationFitness(founderPop, calculateFitnessTwoTrait)

fname <- file.path(save_dir, "3dpopulationfitness.html")
htmlwidgets::saveWidget(as_widget(p), fname)
```

Overlay an adaptive walk over a fitness landscape, and save 3D and 2D versions
```{r}
fig <- plot_ly()
fit_df <- data.frame(gen=1:n.gens,
                   fitness=numeric(n.gens),
                   traitValA=numeric(n.gens),
                   traitValB=numeric(n.gens))

pop <- founderPop

for(gen in 1:n.gens) {
  fit_df$fitness[gen] <- mean(twoTraitFitFunc(gv(pop)))
  fit_df$traitValA[gen] <- meanG(pop)[1]
  fit_df$traitValB[gen] <- meanG(pop)[2]
  meanFitness <- mean(twoTraitFitFunc(pheno(pop)))
  selRat <- selectionRatio(meanFitness)
  pop <- selectCross(pop, trait=twoTraitFitFunc, nInd=n.popSize*selRat, nCrosses=n.popSize)
}

save_dir <- file.path(output_dir, "OverlayAdaptiveWalk")
if (!dir.exists(save_dir)) dir.create(save_dir)
save_dir <- file.path(save_dir, format(Sys.time(), "%F_%H_%M_%S"))
if (!dir.exists(save_dir)) dir.create(save_dir)

plotType <- "CONTOUR"
pc <- overlayWalkOnLandscape(fit_df, type=plotType, calculateFitnessTwoTrait)
fname <- file.path(save_dir, paste0("adaptivewalk_", plotType, ".html"))
htmlwidgets::saveWidget(as_widget(pc), fname)

plotType <- "SURFACE"
ps <- overlayWalkOnLandscape(fit_df, type=plotType, calculateFitnessTwoTrait)
fname <- file.path(save_dir, paste0("adaptivewalk_", plotType, ".html"))
htmlwidgets::saveWidget(as_widget(ps), fname)

write.table(getParams(), file.path(save_dir, "params.txt"), col.names=FALSE, quote=FALSE, sep=":\t")
```

This block will simulate one base population, from which two sub-populations are selected, and undergo purifying selection independently.
```{r}
source("Scripts/GlobalParameters.R")
source("Scripts/CreateFounderPop.R")

fit_df <- data.frame(gen=1:n.burnInGens,
                 fitness=numeric(n.burnInGens),
                 traitValA=numeric(n.burnInGens),
                 traitValB=numeric(n.burnInGens))
pop <- founderPop

# Burn-in generations
for (gen in 1:n.burnInGens) {
  fit_df$fitness[gen] <- mean(twoTraitFitFunc(pheno(pop)))
  fit_df$traitValA[gen] <- meanP(pop)[1]
  fit_df$traitValB[gen] <- meanP(pop)[2]
  pop <- selectCross(pop, trait=twoTraitFitFunc, nInd=nInd(pop)*n.burnInSelProp, nCrosses=nInd(pop))
}

# Create a random vector of size n.pops, with a random order of sub-population ids
randVec <- sample(rep(c(1:n.nPops), times=n.popSize/n.nPops))

# Select all of the "1" indexed individuals
popA <- selectInd(pop, trait=selectSubPop, selectTop=TRUE, nInd=n.subPopSize, idx=1, randVec=randVec)
# Select all of the "2" indexed individuals
popB <- selectInd(pop, trait=selectSubPop, selectTop=TRUE, nInd=n.subPopSize, idx=2, randVec=randVec)

# Create dataframes for each subpopulation, initializing with current values
popA_df <- data.frame(gen=c(1),
                 fitness=c(mean(twoTraitFitFunc(pheno(popA)))),
                 traitValA=c(meanP(popA)[1]),
                 traitValB=c(meanP(popA)[2]))
popB_df <- data.frame(gen=c(1),
                 fitness=c(mean(twoTraitFitFunc(pheno(popB)))),
                 traitValA=c(meanP(popB)[1]),
                 traitValB=c(meanP(popB)[2]))

# Iterate through the generations
for (gen in 1:n.gens) {
  # If popA is within the margin of the fitness optimum, don't progress it any further
  if (mean(twoTraitFitFunc(pheno(popA))) < n.margin) {
    # Advance the population
    meanFitness <- mean(twoTraitFitFunc(pheno(popA)))
    # Get a selection ratio based on fitness
    selRat <- selectionRatio(meanFitness)
    popA <- selectCross(popA, trait=twoTraitFitFunc, nInd=nInd(popA)*selRat, nCrosses=nInd(popA))
    # Update the dataframe with new values
    popA_df <- rbind(popA_df, data.frame(gen=gen,
                                       fitness=meanFitness,
                                       traitValA=meanP(popA)[1],
                                       traitValB=meanP(popA)[2]))
    
  }
  # If popB is within the margin of the fitness optimum, don't progress it any further
  if (mean(twoTraitFitFunc(pheno(popB))) < n.margin) {
    # If popA is within the margin of the fitness optimum, don't progress it any further
    meanFitness <- mean(twoTraitFitFunc(pheno(popB)))
    # Get a selection ratio based on fitnes
    selRat <- selectionRatio(meanFitness)
    popB <- selectCross(popB, trait=twoTraitFitFunc, nInd=nInd(popB)*selRat, nCrosses=nInd(popB))
    # Update the dataframe with new values
    popB_df <- rbind(popB_df, data.frame(gen=gen,
                                       fitness=meanFitness,
                                       traitValA=meanP(popB)[1],
                                       traitValB=meanP(popB)[2]))
  }
}
# Update rownames
rownames(popA_df) <- 1:nrow(popA_df)
rownames(popB_df) <- 1:nrow(popB_df)

# Plot the adaptive walks
fig <- plot_ly()
fig <- add_trace(
  fig,
  fit_df,
  name = "Burn In",
  x = fit_df$traitValA,
  y = fit_df$traitValB,
  z = fit_df$fitness,
  type = 'scatter3d',
  mode = 'lines',
  opacity = 1,
  color = 'yellow',
  line = list(width = 5)
)

fig <- add_trace(
    fig,
    popA_df,
    name = "Pop A",
    x = popA_df$traitValA,
    y = popA_df$traitValB,
    z = popA_df$fitness,
    type = 'scatter3d',
    mode = 'lines',
    opacity = 1,
    color = 'red',
    line = list(width = 5)
  )

fig <- add_trace(
    fig,
    popB_df,
    name = "Pop B",
    x = popB_df$traitValA,
    y = popB_df$traitValB,
    z = popB_df$fitness,
    type = 'scatter3d',
    mode = 'lines',
    opacity = 1,
    color = 'blue',
    line = list(width = 5)
  )


p <- fig %>%
  layout(legend=list(title=list(text='Population')),
         showlegend=FALSE,
         scene = list(xaxis = list(title = "Trait A"),
                      yaxis = list(title = "Trait B"),
                      zaxis = list(title = "Fitness"),
                      aspectmode='cube')) %>% hide_colorbar()

save_dir <- file.path(output_dir, "DivergingPopulations")
if (!dir.exists(save_dir)) dir.create(save_dir)
save_dir <- file.path(save_dir, format(Sys.time(), "%F_%H_%M_%S"))
if (!dir.exists(save_dir)) dir.create(save_dir)
fname <- file.path(save_dir, "adaptivewalks.html")
htmlwidgets::saveWidget(as_widget(p), fname)

fname <- file.path(save_dir, "2PopulationFitness.html")
p <- plot3dPopulationFitnessTwoPops(popA, popB)
htmlwidgets::saveWidget(as_widget(p), fname)
write.table(getParams(), file.path(save_dir, "params.txt"), col.names=FALSE, quote=FALSE, sep=":\t")

fname <- file.path(save_dir, "traitarchitecture.pdf")
pdf(fname)

p1 <- plotTraitArchitecture(popA, "Fitness", "popA")
p2 <- plotTraitArchitecture(popB, "Fitness", "popB")
p3 <- plotTraitArchitecture(popA, "Additive", "popA")
p4 <- plotTraitArchitecture(popB, "Additive", "popB")

(p1|p2)/(p3|p4)
dev.off()

# Create a density plot of trait 1
trait1.df <- as.data.frame(cbind(pheno(popA)[,1], pheno(popB)[,1]))
colnames(trait1.df) <- c("popA", "popB")
trait1.df <- trait1.df %>%
  pivot_longer(c("popA", "popB"), names_to="pop", values_to="pheno")
t1 <- ggplot(trait1.df, aes(pheno, fill=pop, color=pop)) +
  geom_density(alpha=0.1) +
  labs(title="Trait 1")

# Create a density plot of trait 2
trait2.df <- as.data.frame(cbind(pheno(popA)[,2], pheno(popB)[,2]))
colnames(trait2.df) <- c("popA", "popB")
trait2.df <- trait2.df %>%
  pivot_longer(c("popA", "popB"), names_to="pop", values_to="pheno")
t2 <- ggplot(trait2.df, aes(pheno, fill=pop, color=pop)) +
  geom_density(alpha=0.1) +
  labs(title="Trait 2")


(t1|t2)
ggplot2::ggsave(filename = "trait_distributions.pdf",
                path=save_dir,
                device = "pdf")


```

Plot the change in allele frequency over time
```{r}
n.qtlPerChr <- 2
source("Scripts/CreateFounderPop.R")
pop <- founderPop

# Results dataframe
freq.df <- data.frame(gen=1:n.gens)

# Get the effect sizes of each qtl
qtlEff.df <- getQtlEffectSizes(pop)

# Get the names of all the QTLs
qtl <- colnames(getUniqueQtl(pop))

# Create a dataframe of all zeros where the columns are the QTL ids, and the # rows is the # of generations
qtl.df <- data.frame(matrix(0, ncol=length(qtl), nrow=n.gens))
colnames(qtl.df) <- qtl

# Combine the dataframes
freq.df <- cbind(freq.df, qtl.df)

# Iterate through each generation
for(gen in 1:n.gens) {
  meanFitness <- mean(twoTraitFitFunc(pheno(pop)))
  # Get the qtl genotype data
  qtlGeno <- getUniqueQtl(pop)
  # Get the frequency of the '2' allele at each locus
  for (l in 1:length(qtl)) {
    # id is the name of the qtl (chr_site)
    id <- qtl[l]
    # A list of genotype data for each individual in the population at that locus
    locus <- qtlGeno[,id]
    # Calculate the allele frequency as the frequency of homozygous individuals (for 'allele') +
    # 1/2 * frequency of heterozygous individuals (assumes the locus is biallelic)
    freq.df[gen,id] <- (sum(locus==n.allele)/n.popSize) + ((sum(locus==1)/n.popSize)/2)
  }
  # Determine selection ration based on fitness
  selRat <- selectionRatio(meanFitness)
  #Advance individuals
  pop <- selectCross(pop, trait=twoTraitFitFunc, nInd=n.popSize*selRat, nCrosses=n.popSize)
}

# Make the dataframe tidy
freq.df<- melt(freq.df, id="gen", variable.name="QTL ID", value.name="Allele Frequency")

# Add the qtl effect size data to the dataframe
freq.df <- merge(freq.df, qtlEff.df, by.x="QTL ID", by.y="row.names", all.x=TRUE)

# Create a line plot for the change in frequency of alleles over time
# Each line's opacity is a function of its effect size (higher=darker)
g <- ggplot(freq.df, aes(x=gen, y=freq, color=id, alpha=eff_size)) +
  geom_line(size=0.7)

save_dir <- file.path(output_dir, "AlleleFrequencies")
if (!dir.exists(save_dir)) dir.create(save_dir)
save_dir <- file.path(save_dir, format(Sys.time(), "%F_%H_%M_%S"))
if (!dir.exists(save_dir)) dir.create(save_dir)
write.table(getParams(), file.path(save_dir, "params.txt"), col.names=FALSE, quote=FALSE, sep=":\t")

ggplot2::ggsave(filename = paste0("allelefrequencies.pdf"),
                path=save_dir,
                device = "pdf",
                width=10,
                height=7)
```

Simulate an adaptive walk, and plot the following curve:
#segregating QTL/#segregating loci

Original question: out of segregating alleles left in the population, how many
of them would increase fitness? Is this even a valid question?

TODO:
- How does this change with size of allele?
- How does size of 'fitness delta' change over generations?
- Simualate mutations for a single individual to figure out P(allelic substitution is favorable),
should reflect FKO
- P(mutation gets fixed) - show that this matches Kimura
- Compare adaptive walks for DH/inbred population where there are new mutations VS
population w/standing genetic variation
```{r}
n.gens <- 200

# Reset variables
source("Scripts/CreateFounderPop.R")
source("Scripts/GlobalParameters.R")

pop <- founderPop
df <- data.frame(gen=c(),
                 fitness=c(),
                 traitVal1=c(),
                 traitVal2=c(),
                 percFitInc=c(),
                 segAlleles=c(),
                 segQtl=c())

# Iterate through the generations
for (gen in 1:n.gens) {
  print(gen)
  # Counter variables
  numHetLoci = 0
  numHetQtl = 0
  numInc = 0

  # Get all of the loci from the population
  segSiteGeno <- pullSegSiteGeno(pop)
  # Get all of the QTL from the population (a sampling of the segSiteGeno)
  qtlGeno <- getUniqueQtl(pop)
  qtlLoci <- colnames(qtlGeno)
  loci <- colnames(segSiteGeno)
  nLoci <- ncol(segSiteGeno)
  # Iterate through all of the loci
  # TODO: just calculate numHetQtl / numHetLoci
  for (l in 1:ncol(segSiteGeno)) {
    id <- loci[l]
    locus = segSiteGeno[,l]
    # Check if the locus is segregating
    if (hetLocus(locus)) {
      # Increment the counter
      numHetLoci <- numHetLoci + 1
      # Determine the effect size - this is where the bug is
      #e <- getEffectSize(locus, id, pop, "Fitness")
      #if (e > 0) {
      #  numInc <- numInc +1
      #}
      # If the locus is a QTL, then it has an effect size
      if (id %in% qtlLoci) {
        numHetQtl <- numHetQtl + 1
      }
    }
  }

  # Calculate the ratio of segregating QTL to segregating alleles
  if (numHetLoci == 0) {
    perc <- 0
  } else {
    perc <- (numHetQtl / numHetLoci)
  }
  df <- rbind(df, data.frame(gen=gen,
                             fitness=mean(twoTraitFitFunc(gv(pop))),
                             traitVal1=meanG(pop)[1],
                             traitVal2=meanG(pop)[2],
                             percFitInc=perc,
                             segAlleles=numHetLoci,
                             segQtl=numHetQtl))
  # If the population is within n.margin, terminate the simulation
  if (mean(twoTraitFitFunc(gv(pop))) >= n.margin) {
    break
  }
  meanFitness <- mean(twoTraitFitFunc(pheno(pop)))
  selRat <- selectionRatio(meanFitness)
  # Advance the top progeny according to the fitness function
  pop <- selectCross(pop=pop, trait=twoTraitFitFunc, nInd=nInd(pop)*selRat, nCrosses=nInd(pop))
}

save_dir <- file.path(output_dir, "BeneficialAlleles")
if (!dir.exists(save_dir)) dir.create(save_dir)
save_dir <- file.path(save_dir, format(Sys.time(), "%F_%H_%M_%S"))
if (!dir.exists(save_dir)) dir.create(save_dir)
write.table(getParams(), file.path(save_dir, "params.txt"), col.names=FALSE, quote=FALSE, sep=":\t")

fname <- file.path(save_dir, "beneficialalleles.pdf")
pdf(fname)

# Plot the ratio of segregating QTL to segregating alleles, along with fitness
par(mar = c(5, 5, 3, 5) + 0.3)
plot(df$gen, df$fitness, type="l", lwd = "3", col=2, xlab = "Generation", ylab = "Fitness")
par(new = TRUE) 
plot(df$gen, df$percFitInc, type="l", lwd = "3", col = 3, axes = FALSE, xlab = "", ylab = "") 
axis(side = 4, at = pretty(range(df$percFitInc)))
mtext("% of Segregating alleles that are QTL", side = 4, line = 3)
par(xpd=TRUE)
legend("right",
  c("Fitness", "% Alleles"),
       lty = 1,
       lwd = 3,
       col = 2:3)
dev.off()
```
